Linear analysis of the vertical shear instability in astrophysical disks

### 2015-01-23 application.tex
rough version of applications section done

\section{Application to protoplanetary disks}\label{application}

\subsection{Numerical example}

\subsection{Connection with radiative cooling}
As an application of our results, we estimate where in a 
protoplanetary disk do the thermodynamic conditions allow the VSI to 
operate. Specifically, we consider the fundamental isothermal VSI, 
for which we can use the simple criterion Eq. \ref{iso_vsi_cond}. The
condition derived below may be considered as a sufficient condition for the VSI.

@@

\subsubsection{Newtonian cooling}
Our aim is to extract from Eq. \ref{real_energy} the dimensionless
cooling time $\beta$ as used in our linear calculations. To do so, we
consider perturbations with small radial lengthscales $L\equiv
2\pi/k_x<l_\mathrm{rad}$ where 
\begin{align}
l_\mathrm{rad} \equiv \frac{1}{\kappa_d\rho} 
\end{align} 
is the photon mean-free-path and $\kappa_d$ is the (dust) opacity. For
$L< l_\mathrm{rad}$ the radiative cooling is given by {\bf (reference?)}
\begin{align}
\nabla\cdot\bm{F} = 4 \rho \kappa_d \sigma_s T^4,
\end{align}   
where $\sigma_s$ is the Stefan-Boltzmann constant. We choose the
heating term 
\begin{align}
  Q_+ = 4\rho\kappa_d\sigma_s T^{4}_{t=0}. 
\end{align}
Then Eq. \ref{real_energy} becomes
\begin{align}
  \frac{DP}{Dt} = -\gamma P \nabla\cdot\bm{v} -
  \frac{4P\kappa_d\sigma_s}{Tc_v}\left(T^4 - T_{t=0}^4\right). 
\end{align}
When linearized, the last term on the right hand side becomes
\begin{align}
  -\frac{16\sigma_s}{c_v}\kappa_d T^3\left(\delta P -
    \frac{P}{\rho}\delta\rho\right). 
\end{align}
By comparing this expression with the linearized version of the energy
equation used in our calculations (Eq. \ref{energy_eq}), we identify 
\begin{align}
  t_c = \frac{c_v}{16\sigma_s\kappa_dT^3}.
\end{align}
For the fundamental isothermal VSI to operate, we need $\beta =
t_c\Omega_k< \beta_\mathrm{crit}$ as given by
Eq. \ref{iso_vsi_cond}. This translates to
\begin{align}\label{newton_vsi_app} 
  \frac{\mathcal{R}\Omega_k}{16\mu\sigma_s\kappa_dT^3} < \epsilon|q|. 
\end{align}

@@

\subsubsection{Radiative diffusion}
For perturbation lengthscales $L>l_\mathrm{rad}$,  
let us assume that temperature fluctuations are smoothed out by
radiative diffusion. To model this in linear theory we can redefine
the cooling time as the operator 
\begin{align}
  t_c^{-1} \to -\eta \nabla^2, 
\end{align}
where $\eta$ is the diffusion coefficient. This, of course,
fundamentally changes the linear problem by introducing higher order
derivatives. We defer a proper analysis of this problem to a follow-up
study. Here, we are interested in order-of-magnitude estimates. We
thus proceed by assuming vertical derivatives can be neglected in the
Laplacian. Then
\begin{align}
  t_c^{-1} \sim \eta k_x^2 = \hat{\eta} \Omega_k \khat^2,
\end{align}
where $\hat{\eta} = \eta/H_\mathrm{iso}^2\Omega_k$ is the
dimensionless diffusion coefficient, so that
\begin{align}
  \beta = \frac{1}{\hat{\eta}\khat^2}. 
\end{align}

@@

Now, the dimensional diffusion cofficient is {\bf(reference needed?)}
\begin{align}
  \eta = \frac{16\sigma_s T^3}{3\kappa_d\rho^2 c_v}, 
\end{align}
so that
\begin{align}
  \hat{\eta} \simeq  \frac{16\sigma_s T^3}{3\kappa_d\Sigma^2
    c_v\Omega_k}. 
\end{align}
The criterion Eq. \ref{iso_vsi_cond_diff} becomes
\begin{align}
\frac{3\kappa_d\Sigma^2\mathcal{R}
  \Omega_k}{16\mu\sigma_s T^3} < \epsilon |q| \khat^2. 
\end{align}
For the fiducial disk model this translates to
\begin{align}
  \left(\frac{r}{\mathrm{AU}}\right) \gtrsim
  76.5\left(\frac{\hat{\kappa}_d\hat{\Sigma}^2}{\hat{T}}\right)^{14/57}\mu^{-14/57}\khat^{-28/57}. 
\end{align} 
Inserting $\mu=2$ and $\khat=10$ we find, for the MMSN, that
\begin{align*}
  r\gtrsim 21 \mathrm{AU}. 
\end{align*}

@@

>>>>

\section{Application to protoplanetary disks}\label{application} 
As an application of our results, we estimate where in a 
protoplanetary disk do the thermodynamic conditions allow the VSI to 
operate. Specifically, we consider the fundamental isothermal VSI, 
for which we can use the simple criterion Eq. \ref{iso_vsi_cond}. The
condition derived below may be considered as a sufficient condition for the VSI.

@@

\subsection{Newtonian cooling}
Our aim is to extract from Eq. \ref{real_energy} the dimensionless
cooling time $\beta$ as used in our linear calculations. To do so, we
consider perturbations with small radial lengthscales $L\equiv
2\pi/k_x<l_\mathrm{rad}$ where 
\begin{align}
l_\mathrm{rad} \equiv \frac{1}{\kappa_d\rho} 
\end{align} 
is the photon mean-free-path and $\kappa_d$ is the (dust) opacity. For
$L< l_\mathrm{rad}$ the radiative cooling is given by {\bf (reference?)}
\begin{align}
\nabla\cdot\bm{F} = 4 \rho \kappa_d \sigma_s T^4,
\end{align}   
where $\sigma_s$ is the Stefan-Boltzmann constant. We choose the
heating term 
\begin{align}
  Q_+ = 4\rho\kappa_d\sigma_s T^{4}_{t=0}. 
\end{align}
Then Eq. \ref{real_energy} becomes
\begin{align}
  \frac{DP}{Dt} = -\gamma P \nabla\cdot\bm{v} -
  \frac{4P\kappa_d\sigma_s}{Tc_v}\left(T^4 - T_{t=0}^4\right). 
\end{align}
When linearized, the last term on the right hand side becomes
\begin{align}
  -\frac{16\sigma_s}{c_v}\kappa_d T^3\left(\delta P -
    \frac{P}{\rho}\delta\rho\right). 
\end{align}
By comparing this expression with the linearized version of the energy
equation used in our calculations (Eq. \ref{energy_eq}), we identify 
\begin{align}\label{tc_newton_cool}
  t_c = \frac{c_v}{16\sigma_s\kappa_dT^3}.
\end{align}
For the fundamental isothermal VSI to operate, we need $\beta =
t_c\Omega_k< \beta_\mathrm{crit}$ as given by
Eq. \ref{iso_vsi_cond}. This translates to
\begin{align}\label{newton_vsi_app} 
  \frac{\mathcal{R}\Omega_k}{16\mu\sigma_s\kappa_dT^3} < \epsilon|q|. 
\end{align}

@@

\subsubsection{Numerical example 1}
Consider the MMSN at $r=50\mathrm{AU}$, with $\gamma=1.4$ and
$\mu=2$. Then we find
the dimensionless thermal relaxation time, using
Eq. \ref{tc_newton_cool}, is
\begin{align*}
  \beta \simeq 0.057. 
\end{align*}
At $r=50\mathrm{AU}$, the disk aspect ratio is $\epsilon \simeq
0.067$ which implies, using Eq. \ref{iso_vsi_cond}, a critical 
relaxation timescale of  
\begin{align*}
  \beta_\mathrm{crit} \simeq 0.072. 
\end{align*}
Thus, the disk supports the fundamental isothermal VSI, but only
marginally. Using the above value of $\beta$ and the disk parameters
for the MMSN at $50\mathrm{AU}$ in our linear code we find, for a
perturbation wavenumber $\khat=10$, a growth rate of
\begin{align*}
  \nu = 2.6\times10^{-3}\Omega_k.
\end{align*}
This corresponds to a characteristic growth time of $t_g\simeq 60$
orbits. 
%only find one unstable mode -> close to marginal

\subsection{Radiative diffusion}
For perturbation lengthscales $L>l_\mathrm{rad}$,  
let us assume that temperature fluctuations are smoothed out by
radiative diffusion. To model this in linear theory we can redefine
the cooling time as the operator 
\begin{align}
  t_c^{-1} \to -\eta \nabla^2, 
\end{align}
where $\eta$ is the diffusion coefficient. This, of course,
fundamentally changes the linear problem by introducing higher order
derivatives. We defer a proper analysis of this problem to a follow-up
study. Here, we are interested in order-of-magnitude estimates. We
thus proceed by assuming vertical derivatives can be neglected in the
Laplacian. Then
\begin{align}
  t_c^{-1} \sim \eta k_x^2 = \hat{\eta} \Omega_k \khat^2,
\end{align}
where $\hat{\eta} = \eta/H_\mathrm{iso}^2\Omega_k$ is the
dimensionless diffusion coefficient, so that
\begin{align}
  \beta = \frac{1}{\hat{\eta}\khat^2}. 
\end{align}

@@

Now, the dimensional diffusion cofficient is {\bf(reference needed?)}
\begin{align}
  \eta = \frac{16\sigma_s T^3}{3\kappa_d\rho^2 c_v}, 
\end{align}
so that
\begin{align}\label{diff_coeff_dimensionless}
  \hat{\eta} \simeq  \frac{16\sigma_s T^3}{3\kappa_d\Sigma^2
    c_v\Omega_k}. 
\end{align}
The criterion Eq. \ref{iso_vsi_cond_diff} becomes
\begin{align}
\frac{3\kappa_d\Sigma^2\mathcal{R}
  \Omega_k}{16\mu\sigma_s T^3} < \epsilon |q| \khat^2. 
\end{align}
For the fiducial disk model this translates to
\begin{align}
  \left(\frac{r}{\mathrm{AU}}\right) \gtrsim
  76.5\left(\frac{\hat{\kappa}_d\hat{\Sigma}^2}{\hat{T}}\right)^{14/57}\mu^{-14/57}\khat^{-28/57}. 
\end{align} 
Inserting $\mu=2$ and $\khat=10$ we find, for the MMSN, that
\begin{align*}
  r\gtrsim 21 \mathrm{AU}. 
\end{align*}

@@

\subsubsection{Numerical example 2}
Consider the MMSN at $r=25\mathrm{AU}$, with $\gamma=1.4$ and $\mu=2$
as before. Evaluating Eq. \ref{diff_coeff_dimensionless} at
$25\mathrm{AU}$ for the MMSN, we find
\begin{align*}
  \frac{1}{\hat{\eta}} \simeq 1.12. 
\end{align*}
At $r=25\mathrm{AU}$ the disk aspect ratio is
$\epsilon\simeq0.055$. Then Eq. \ref{iso_vsi_cond_diff} becomes
\begin{align*}
  \khat^2\gtrsim 15. 
\end{align*}

Using the above disk parameters with $\beta =
1/\hat{\eta}\khat^2\simeq0.011$ in our linear code, we find, for
$\khat=10$ a growth rate of 
\begin{align*}
\nu = 8.5\times10^{-3}\Omega_k. 
\end{align*} 
This corresponds to a characteristic growth time of $t_g\simeq19$
orbits. 
%other modes exist but fundamental isothermal is among the fastest
%(second harmonic only slightly faster)
